---
title: "MAPLE Report"
title-block-banner: true
title-block-banner-color: '#CC4945'
subtitle: "`r format(Sys.Date(),format=c('%d-%b-%Y'))`"
format: 
  html:
    self-contained: true
    page-layout: full
    theme: cosmo
    toc: false
    toc_float: false
editor: visual
execute: 
  warning: false
  echo: false
params:
  runid: "Default"
  working_dir: "."
  marker: "Unknown"
  ampsize: 0
  numreps: 0
---

```{css,echo=FALSE}
.title {
  font-size: 50px;
  color: #f1f1f1;
  font-family: Arial;
}

.section-box {
  background-color: #f8f9fa;
  padding: 20px;
  border-radius: 8px;
  box-shadow: inset 0 2px 5px rgba(0,0,0,0.05);
  margin-bottom: 20px;
}

.text-center {
  text-align: center;
}

.scrollme {
  max-height: 140px;
  overflow-y: auto;
  overflow-x: hidden;
  padding-right: 12px;
}
```

```{r}
#| label: setup
#| include: true
#| echo: false
#| warning: false

# Import aarguments
runid <- params$runid
wkdir <- params$working_dir
marker <- params$marker
ampsize <- params$ampsize
num.reps <- params$numreps

# Required packages
packages <- c(
  "dplyr", "tidyr", "plyr", "readxl", "leaflet", "plotly", "forcats",
  "shiny", "iNEXT", "kableExtra", "treemapify", "paletteer", "htmltools", "RColorBrewer",
  "stringr")

# Install any packages that are not already installed
installed_packages <- packages %in% rownames(installed.packages())
if(any(!installed_packages)) {
  install.packages(packages[!installed_packages])
}

# Load all packages
invisible(lapply(packages, 
                 library, 
                 character.only = TRUE))
```

```{r}
#| include: true
#| echo: false

setwd(wkdir)

# Paths for metabarcoding output and the parameter file. 
metabarcoding_output <- sprintf("Metabarcoding Results - %s_%s_%s.xlsx", runid, marker, ampsize)

parameters <- list.files(pattern = "parameters*", path = "../", full.names = TRUE)

# Import data by replicates
metabarcode_data_by_replicate<-read_excel(metabarcoding_output, sheet=2)

# Export as tsv
metabarcode_data_by_replicate_tsv <- metabarcode_data_by_replicate
names(metabarcode_data_by_replicate_tsv) <- gsub(" ", "_", names(metabarcode_data_by_replicate_tsv))

write.table(metabarcode_data_by_replicate_tsv, sprintf("Metabarcoding_Results_%s_%s_%s_ByReplicate.tsv", runid, marker, ampsize), quote = F, row.names = F, sep = "\t")

# Import data by sample
metabarcode_data_by_sample<-read_excel(metabarcoding_output, sheet=1)

# Export as tsv
metabarcode_data_by_sample_tsv <- metabarcode_data_by_sample
names(metabarcode_data_by_sample_tsv) <- gsub(" ", "_", names(metabarcode_data_by_sample_tsv))

write.table(metabarcode_data_by_sample_tsv, sprintf("Metabarcoding_Results_%s_%s_%s_BySample.tsv", runid, marker, ampsize), quote = F, row.names = F, sep = "\t")

# Parameters metadata for mapping and sampling summary
mapping <- read.delim(list.files(pattern = "temp_mapping*"), header = F, check.names = F, fill = T)

metadata <- read_excel(parameters, sheet = 3) %>%
  {
    date_cols <- c("Collection Start Date (D-Mmm-YY)", "Collection End Date (D-Mmm-YY)")
    
    # Check if both columns exist and contain at least one non-NA value
    if (all(date_cols %in% names(.)) && any(!is.na(unlist(select(., all_of(date_cols)))))) {
      dplyr::mutate(., across(all_of(date_cols), ~ suppressWarnings(as.Date(.x, format = "%d-%b-%y"))))
    } else {
      .  # Skip mutate if columns missing or empty
    }
  }

# Combined data to get replicate information along with the BIN assignment
metabarcode_comb_data <- metabarcode_data_by_replicate%>%
    dplyr::select(Sample, Reads, Replicate,ASV_ID, Kingdom:Species)%>%left_join(
    metabarcode_data_by_sample%>%
      select(Sample,ASV_ID, `BIN Match Status`, `BIN Hit`),
    by=c('Sample','ASV_ID'))

# All samples
sample.list <- mapping %>%
  # Remove trailing _Rep<number>
  mutate(sample.list = str_remove(V3, "_Rep\\d+$")) %>%
  # Filter out unwanted sample names
  filter(!str_detect(sample.list, regex("NEG|NEGATIVE|CONTROL|BLANK", ignore_case = TRUE))) %>%
  # Keep unique
  distinct(sample.list) %>%
  pull(sample.list)


# Samples with data
sample.w.data.list <- unique(metabarcode_data_by_sample$Sample)

# Number of samples with data
sample.nos.w.data<-length(sample.w.data.list)
```

# Run Summary {.section-box .text-center}

```{r}
#| echo: false

Run_ID <- runid

Marker <- marker

Amplicon_Size <- sprintf("%s bp", ampsize)

Number_of_samples<-length(sample.list)

Number_of_Technical_Replicates <- num.reps

Number_of_samples_with_GPS_coordinates <- metadata %>%
  filter(!is.na(Latitude) & !is.na(Longitude)) %>%
  summarise(n = n_distinct(Sample)) %>%
  pull(n)

Number_of_samples_with_collection_dates <- metadata %>%
  filter(if_any(starts_with("Collection"), ~ !is.na(.))) %>%
  summarise(n = n_distinct(Sample)) %>%
  pull(n)

Total_unique_BINs_detected<-metabarcode_data_by_sample%>%
    filter(`BIN Match Status`=='BIN_MATCH')%>%  
    dplyr::summarise(Unique_BINs = n_distinct(`BIN Hit`))

Mean_BINs_per_sample<-metabarcode_comb_data%>%
  filter(`BIN Match Status`=='BIN_MATCH')%>% 
  group_by(Sample)%>%
  dplyr::summarise(bin_counts=n_distinct(`BIN Hit`))%>%
  dplyr::summarise(Mean_BINs_per_sample=round(mean(bin_counts,na.rm=T),0))

Max_BINs_per_sample<-metabarcode_comb_data%>%
  filter(`BIN Match Status`=='BIN_MATCH')%>% 
  group_by(Sample)%>%
  dplyr::summarise(bin_counts=n_distinct(`BIN Hit`))%>%
  dplyr::summarise(Max_BINs_per_sample=max(bin_counts,na.rm=T))

Min_BINs_per_sample<-metabarcode_comb_data%>%
  filter(`BIN Match Status`=='BIN_MATCH')%>% 
  group_by(Sample)%>%
  dplyr::summarise(bin_counts=n_distinct(`BIN Hit`))%>%
  dplyr::summarise(Min_BINs_per_sample=min(bin_counts,na.rm=T))

collection_period <- metadata %>%
  {
    start_col <- "Collection Start Date"
    end_col   <- "Collection End Date"
    
    # Check if both date columns exist and contain any non-NA values
    if (all(c(start_col, end_col) %in% names(.)) &&
        any(!is.na(.[[start_col]])) &&
        any(!is.na(.[[end_col]]))) {
      
      summarise(.,
                collection_period_start = suppressWarnings(min(as.Date(.[[start_col]]), na.rm = TRUE)),
                collection_period_end   = suppressWarnings(max(as.Date(.[[end_col]]), na.rm = TRUE))) %>%
        mutate(collection_period = ifelse(
          is.na(collection_period_start) | is.na(collection_period_end),
          NA,
          paste(collection_period_start, "to", collection_period_end)
        )) %>%
        pull(collection_period)
      
    } else {
      NA_character_  # If missing or empty, return NA cleanly
    }
  }

summary_table <- data.frame(
  Run_ID,
  Marker,
  Amplicon_Size,
  Number_of_samples,
  Number_of_Technical_Replicates,
  Number_of_samples_with_GPS_coordinates,
  Number_of_samples_with_collection_dates,
  Collection_period = collection_period,
  Total_unique_BINs_detected,
  Mean_BINs_per_sample,
  Max_BINs_per_sample,
  Min_BINs_per_sample
) %>%
  mutate(across(everything(), as.character)) %>%
  rename_with(~ gsub("_", " ", .x)) %>%
  pivot_longer(
    cols = everything(),
    names_to = "Category",
    values_to = "Value"
  )
```

```{r}
#| echo: false

# Generating the interactive summary table
summary_table|>
  kbl(col.names = NULL) %>%
   kable_styling(bootstrap_options = c("hover", "condensed", "responsive"))
```

# Map of sampling locations {.section-box .text-center}

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 6

if (!("Latitude" %in% names(metadata)) ||
    !("Longitude" %in% names(metadata)) ||
    all(is.na(metadata$Latitude)) ||
    all(is.na(metadata$Longitude))) {

  # No coordinates: show a styled HTML message box instead of a map
  htmltools::HTML('
    <div style="
      margin: 30px auto;
      padding: 20px;
      max-width: 500px;
      background-color: #fff3cd;
      border: 1px solid #ffeeba;
      border-radius: 10px;
      color: #856404;
      font-size: 16px;
      text-align: center;
      font-weight: bold;
      box-shadow: 0 2px 6px rgba(0,0,0,0.2);
    ">
      ⚠️ No sample coordinates provided.
    </div>
  ')

} else {

  # Make sure Latitude/Longitude are numeric
  metadata <- metadata %>%
    dplyr::mutate(
      Latitude = as.numeric(Latitude),
      Longitude = as.numeric(Longitude)
    ) %>%
    dplyr::filter(!is.na(Latitude) & !is.na(Longitude))

  # Only create markers if at least one valid row exists
  if (nrow(metadata) == 0) {
    htmltools::HTML('
      <div style="
        margin: 30px auto;
        padding: 20px;
        max-width: 500px;
        background-color: #fff3cd;
        border: 1px solid #ffeeba;
        border-radius: 10px;
        color: #856404;
        font-size: 16px;
        text-align: center;
        font-weight: bold;
        box-shadow: 0 2px 6px rgba(0,0,0,0.2);
      ">
        ⚠️ No valid sample coordinates found.
      </div>
    ')
  } else {

    leaflet(metadata) %>%
      addProviderTiles("Esri.WorldImagery", group = "ESRI_WorldImage") %>%
      addProviderTiles("Esri.WorldTopoMap",
                       options = providerTileOptions(noWrap = TRUE),
                       group = "ESRI_WorldTopoMap") %>%
      addCircleMarkers(
        lng = ~Longitude,
        lat = ~Latitude,
        label = ~paste0(
          "<b>Sample:</b> ", Sample, "<br>",
          "<b>Site:</b> ", `Collection Site`, "<br>",
          "<b>Lat:</b> ", round(Latitude, 3), "<br>",
          "<b>Lon:</b> ", round(Longitude, 3)
        ),
        labelOptions = labelOptions(
          textsize = "11px",
          direction = "auto",
          style = list("background" = "white", "padding" = "5px", "border-radius" = "5px")
        ),
        radius = 7,
        opacity = 1,
        color = "black",
        weight = 1.8,
        fillColor = "#CC4945",
        stroke = TRUE,
        fillOpacity = 1
      ) %>%
      addScaleBar(position = "topright") %>%
      addLayersControl(
        baseGroups = c("ESRI_WorldImage", "ESRI_WorldTopoMap"),
        options = layersControlOptions(collapsed = FALSE)
      )
  }
}

```

# Sample metadata {.section-box .text-center}

::: column-page
```{r}
#| echo: false

# Sample information
DT::datatable(
  metadata%>%mutate(Latitude=round(Latitude,3),
                    Longitude=round(Longitude,3)),
  options = list(
    pageLength = 10,
    autoWidth = TRUE,
    scrollX = TRUE
  ),
  filter = 'top',
  class = 'cell-border stripe'
)

```
:::

# Taxonomic Composition of Samples {.section-box .text-center}

*Hover over bars for details. Double-click legend elements to isolate them. Single-click legend elements to select/deselect specific groups. Double-click on legend to reset view.*

:::: column-page
```{r}
################ COMMON GROUPS ##########################
# Create dictionary to match taxonomic names with common name groups
metabarcode_data_by_sample_2 <- metabarcode_data_by_sample %>%
  mutate(
    category = case_when(
      # Microbial groups
      Kingdom == "Bacteria" ~ "Bacteria",
      Kingdom == "Fungi" ~ "Fungi",
      Kingdom == "Protista" ~ "Protists",

      # Nematodes
      Phylum == "Nematoda" ~ "Nematodes",

      # Invertebrate groups
      Class == "Insecta" ~ "Insects",
      Class == "Arachnida" ~ "Arachnids",
      Class == "Mollusca" ~ "Molluscs",

      # Vertebrate groups
      Class %in% c("Actinopterygii", "Chondrichthyes", "Sarcopterygii") ~ "Fish",
      Class == "Aves" ~ "Birds",
      Class == "Mammalia" ~ "Mammals",
      Class %in% c("Reptilia", "Amphibia") ~ "Reptiles/Amphibians",

      # Fallback
      TRUE ~ "Other"
    )
  )

# Summarize sample compositions for common taxonomic groups
sample_composition <- metabarcode_data_by_sample_2 %>%
  group_by(Sample, category) %>%
  summarise(
    n_ASVs = n_distinct(ASV_ID),
    .groups = "drop"
  ) %>%
  group_by(Sample) %>%
  mutate(
    total_ASVs = sum(n_ASVs),
    percent = 100 * n_ASVs / total_ASVs
  ) %>%
  ungroup() %>%
  arrange(Sample, category)

# Create vertical stacked bar plot
p_stack <- ggplot(
  sample_composition,
  aes(
    y = factor(Sample, levels = rev(sort(unique(Sample)))),       
    x = percent,
    fill = category,
    text = paste0(
      "<b>Sample:</b> ", Sample, "<br>",
      "<b>Category:</b> ", category, "<br>",
      "<b>ASVs:</b> ", n_ASVs, "<br>",
      "<b>Percent:</b> ", sprintf("%.1f%%", percent)
    )
  )
) +
  geom_col(width = 0.8) +
  scale_x_continuous(         
    limits = c(0, 100),
    expand = c(0, 0)
  ) +
  labs(
    x = "Percent of ASVs",
    y = NULL,
    fill = "Category"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.y = element_text(size = 8),
    panel.grid.minor = element_blank()
  )

####################### RANKS ##########################

make_taxa_stackplot <- function(df, rank_col, top_n = 8, px_per_sample = 20, min_height = 400) {
  
  rank_col  <- rlang::ensym(rank_col)
  rank_name <- rlang::as_string(rank_col)

  # Replace NA with "unknown" 
  df <- df %>%
    mutate(!!rank_col := ifelse(is.na(!!rank_col), "unknown", !!rank_col))

  # Count ASVs per Sample × Taxon 
  comp <- df %>%
    group_by(Sample, taxon = !!rank_col) %>%
    summarise(n_ASVs = n_distinct(ASV_ID), .groups = "drop") %>%
    mutate(taxon = trimws(taxon))

  # Determine top N taxa based on # of SAMPLES 
  top_taxa <- comp %>%
    filter(taxon != "unknown") %>%
    group_by(taxon) %>%
    summarise(n_samples = n_distinct(Sample), .groups = "drop") %>%
    arrange(desc(n_samples)) %>%
    slice_head(n = top_n) %>%
    pull(taxon)

  # Collapse non-top taxa into "Other" 
  comp2 <- comp %>%
    mutate(taxon = case_when(
      taxon == "unknown" ~ "unknown",
      taxon %in% top_taxa ~ taxon,
      TRUE ~ "Other"
    )) %>%
    group_by(Sample, taxon) %>%
    summarise(n_ASVs = sum(n_ASVs), .groups = "drop") %>%
    group_by(Sample) %>%
    mutate(percent = 100 * n_ASVs / sum(n_ASVs)) %>%
    ungroup()

  comp2$taxon <- trimws(as.character(comp2$taxon))

  # Fixed colour mapping with dynamic extension 
  if (rank_name == "Kingdom") {
    fixed_cols <- c(
      "Animalia" = "#1E88E5",
      "Bacteria" = "#FDD835",
      "Fungi"    = "#8E24AA",
      "Protista" = "#43A047"
    )
  } else if (rank_name == "Phylum") {
    fixed_cols <- c(
      "Arthropoda" = "#1E88E5",
      "Nematoda"   = "#E53935",
      "Mollusca"   = "#8E24AA"
    )
  } else if (rank_name == "Class") {
    fixed_cols <- c(
      "Insecta"   = "#4CAF50",
      "Arachnida" = "#E53935"
    )
  } else if (rank_name == "Order") {
    fixed_cols <- c(
      "Lepidoptera" = "#4CAF50",
      "Diptera"     = "#1E88E5",
      "Coleoptera"  = "#FDD835",
      "Hymenoptera" = "#E53935",
      "Hemiptera"   = "#673AB7",
      "Orthoptera"  = "#8D6E63"
    )
  } else {
    fixed_cols <- c()
  }

  # Add dynamic colours for any remaining taxa
  taxa_present <- setdiff(unique(comp2$taxon), c("Other", "unknown"))
  extra_taxa <- setdiff(taxa_present, names(fixed_cols))
  if (length(extra_taxa) > 0) {
    extra_cols <- colorRampPalette(RColorBrewer::brewer.pal(8, "Set3"))(length(extra_taxa))
    names(extra_cols) <- extra_taxa
    fixed_cols <- c(fixed_cols, extra_cols)
  }

  # Always include Other + unknown
  if (!"Other" %in% names(fixed_cols)) fixed_cols["Other"] <- "gray60"
  if (!"unknown" %in% names(fixed_cols)) fixed_cols["unknown"] <- "black"

  # Factor ordering 
  ordered_main <- intersect(names(fixed_cols), taxa_present)
  ordered_main <- setdiff(ordered_main, c("Other", "unknown"))
  taxa_order <- c(ordered_main, "Other", "unknown")
  taxa_order <- taxa_order[taxa_order %in% unique(comp2$taxon)]
  comp2$taxon <- factor(comp2$taxon, levels = taxa_order)

  # Dynamic height 
  n_samples <- n_distinct(comp2$Sample)
  plot_height <- max(min_height, n_samples * px_per_sample)

  # Generate plot 
  p <- ggplot(comp2, aes(
      y = factor(Sample, levels = rev(sort(unique(Sample)))),
      x = percent,
      fill = taxon,
      text = paste0(
        "<b>Sample:</b> ", Sample, "<br>",
        "<b>", rank_name, ":</b> ", taxon, "<br>",
        "<b>ASVs:</b> ", n_ASVs, "<br>",
        "<b>Percent:</b> ", sprintf("%.1f%%", percent)
      )
  )) +
    geom_col(width = 0.8, color = "black") +
    scale_x_continuous(expand = c(0,0)) +
    scale_fill_manual(values = fixed_cols) +
    labs(x = "Percent of ASVs", y = NULL, fill = rank_name) +
    theme_minimal(base_size = 14) +
    theme(axis.text.y = element_text(size = 8),
          panel.grid.minor = element_blank())

  ggplotly(p, tooltip = "text")
}


```

::: panel-tabset
### Common Groups

```{r}
#| class-output: scrollme
n_samples <- n_distinct(sample_composition$Sample)
px_per_sample <- 20      # Per sample
min_height <- 400        # Minimum height
plot_height_common <- max(min_height, n_samples * px_per_sample)

plotly::ggplotly(p_stack, tooltip = "text") %>%
  layout(height = plot_height_common)

```

### Kingdom

```{r}
#| class-output: scrollme
p_kingdom <- make_taxa_stackplot(metabarcode_data_by_sample_2, Kingdom)
plotly::layout(p_kingdom, height = plot_height_common)

```

### Phylum

```{r}
#| class-output: scrollme
p_phylum <- make_taxa_stackplot(metabarcode_data_by_sample_2, Phylum)
plotly::layout(p_phylum, height = plot_height_common)
```

### Class

```{r}
#| class-output: scrollme
p_class <- make_taxa_stackplot(metabarcode_data_by_sample_2, Class)
plotly::layout(p_class, height = plot_height_common)
```

### Order

```{r}
#| class-output: scrollme
p_order <- make_taxa_stackplot(metabarcode_data_by_sample_2, Order)
plotly::layout(p_order, height = plot_height_common)
```

### Family

```{r}
#| class-output: scrollme
p_family <- make_taxa_stackplot(metabarcode_data_by_sample_2, Family)
plotly::layout(p_family, height = plot_height_common)
```

### Genus

```{r}
#| class-output: scrollme
p_genus <- make_taxa_stackplot(metabarcode_data_by_sample_2, Genus)
plotly::layout(p_genus, height = plot_height_common)
```

### Species

```{r}
#| class-output: scrollme
p_species <- make_taxa_stackplot(metabarcode_data_by_sample_2, Species)
plotly::layout(p_species, height = plot_height_common)
```
:::
::::

# BIN overlap among `r num.reps` replicates {.section-box .text-center}

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 7

# Calculate % overlap between replicates 

calculate_shared_bins <- function(df, num.reps) {
  df %>%
    filter(`BIN Match Status` == "BIN_MATCH") %>%
    dplyr::group_by(Sample) %>%
    dplyr::summarize(
      total_unique_BINs = n_distinct(`BIN Hit`),
      shared_BINs = n_distinct(`BIN Hit`[Replicates == num.reps]),
      percent_shared = ifelse(
        total_unique_BINs > 0,
        (shared_BINs / total_unique_BINs) * 100,
        0
      ),
      .groups = "drop"
    )
}

shared_bin_percentages <- data.frame(calculate_shared_bins(metabarcode_data_by_sample, num.reps))

missing_samples <- setdiff(sample.w.data.list, shared_bin_percentages$Sample)

if (length(missing_samples) != 0) {
  new_rows <- data.frame(
    Sample = missing_samples,
    total_unique_BINs = 0,
    shared_BINs = 0,
    percent_shared = 0
  )
  shared_bin_percentages <- rbind(shared_bin_percentages, new_rows)
}

# Violin plot

plot1_all_asv <- ggplot(shared_bin_percentages, 
                        aes(x = "", 
                            y = percent_shared)) +
  geom_violin(fill = "#0077b6", alpha = 0.6, color = "#023047", trim = FALSE) +
  geom_boxplot(width = 0.1, color = "black", alpha = 0.7, outlier.size = 1) +
  # Add hover text aesthetic
  geom_jitter(
    aes(text = paste0(Sample, " (", round(percent_shared, 0), "%)")),
    width = 0.08,
    alpha = 0.5,
    size = 1.2,
    color = "#011826"
  ) +
  theme_bw(base_size = 14) +
  coord_cartesian(ylim = c(0, 100)) +
  scale_y_continuous(expand = c(0, 0), breaks = seq(0, 100, 10)) +
  labs(y = "% overlap of BINs", x = "", title = "") +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.y = element_blank()
  )

ggplotly(plot1_all_asv, tooltip = "text")

```

# BIN accumulation curves based on `r num.reps` replicates for `r sample.nos.w.data` samples with BINs {.section-box .text-center}

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 10
#| results: asis

library(dplyr)
library(ggplot2)
library(plotly)
library(htmltools)

# Create an empty list to store results 
plot_data_list <- list()

# Iterate over each sample to calculate the average cumulative BINs across combinations 
for (sample in sample.w.data.list) {

  # Subset the data for the current sample
  metabarcode_comb_data_sample <- metabarcode_comb_data %>%
    filter(`BIN Match Status` == "BIN_MATCH", Sample == sample)

  # Initialize a vector to store averaged cumulative BIN counts
  avg_cumulative_bins <- numeric(num.reps)

  # Iterate over replicate counts (k = 1 to num.reps)
  for (k in seq_len(num.reps)) {

    # Generate all combinations of k replicates
    replicate_combinations <- combn(paste0("Rep", 1:num.reps), k, simplify = FALSE)

    # Calculate unique BINs for each combination
    bin_counts <- sapply(replicate_combinations, function(replicate_group) {
      current_data <- metabarcode_comb_data_sample %>%
        filter(Replicate %in% replicate_group)
      length(unique(current_data$`BIN Hit`))
    })

    # Store the average number of unique BINs across all combinations of k replicates
    avg_cumulative_bins[k] <- mean(bin_counts)
  }

  # Create a dataframe for plotting
  plot_data <- data.frame(
    Sample = rep(sample, num.reps),
    Replicate_Count = seq_len(num.reps),
    Avg_Cumulative_BINs = avg_cumulative_bins
  )

  # Add the result to the list
  plot_data_list[[sample]] <- plot_data
}

# Combine all sample data into one data frame for plotting 
plot_data_combined <- bind_rows(plot_data_list)

# Determine Y-axis limits dynamically ---
y_max <- max(plot_data_combined$Avg_Cumulative_BINs)
y_min <- min(plot_data_combined$Avg_Cumulative_BINs)

# Define rounding factor based on the value of y_max 
if (y_max - y_min <= 10) {
  rounding_factor <- 1
} else if (y_max - y_min <= 100) {
  rounding_factor <- 10
} else if (y_max - y_min <= 1000) {
  rounding_factor <- 100
} else {
  rounding_factor <- 1000
}

# Round limits to appropriate factor 
y_upper_limit <- ceiling(y_max / rounding_factor) * rounding_factor
y_lower_limit <- floor(y_min / rounding_factor) * rounding_factor

# Determine facet layout and height 
n_samples <- length(unique(plot_data_combined$Sample))
n_cols <- 4
n_rows <- ceiling(n_samples / n_cols)

facet_height <- 300
internal_height <- n_rows * facet_height

scroll_window_height <- 750

# Create the ggplot 
plot1_accum_bin <- ggplot(plot_data_combined,
                          aes(x = Replicate_Count, y = Avg_Cumulative_BINs)) +
  geom_line(color = "#1EC2A0", linewidth = 1) +
  geom_point(size = 2, color = "#CC4945") +
  scale_x_continuous(breaks = seq(1, num.reps, 1),
                     limits = c(0, num.reps + 0.2)) +
  scale_y_continuous(
    limits = c(y_lower_limit - rounding_factor * 1.1,
               y_upper_limit + rounding_factor),
    breaks = seq(y_lower_limit,
                 y_upper_limit + rounding_factor,
                 rounding_factor)
  ) +
  facet_wrap(~ Sample, ncol = n_cols) +
  theme_bw(base_size = 14) +
  theme(
    axis.line = element_line(color = "#011826"),
    panel.grid.minor.x = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.y = element_blank()
  ) +
  labs(
    x = "Number of Replicates",
    y = "Cumulative BINs"
  ) +
  coord_cartesian(clip = "off", expand = FALSE)

# Convert to Plotly with dynamic height 
p_plotly <- ggplotly(plot1_accum_bin) %>% layout(height = internal_height)

# Wrap in scrollable container 
htmltools::div(
  style = paste0(
    "overflow-y: auto; ",
    "height:", scroll_window_height, "px; ",
    "border: 1px solid #ccc; padding: 5px;"
  ),
  p_plotly
)

```

# ASV BIN matches for each of the `r sample.nos.w.data` samples {.section-box .text-center}

*Hover over points for additional details. Highlight area to zoom. Double click anywhere to return to normal view.*

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
library(plotly)

# BIN counts
bin_count <- metabarcode_comb_data %>%
  filter(`BIN Match Status` == "BIN_MATCH") %>%
  group_by(Sample) %>%
  summarise(Unique_BINs = n_distinct(`BIN Hit`), .groups = "drop")

# Non-BIN counts 
nobin_count <- metabarcode_comb_data %>%
  filter(`BIN Match Status` == "NO_MATCH") %>%
  group_by(Sample) %>%
  summarise(Unique_Unknown_Sequences = n_distinct(`BIN Hit`), .groups = "drop")

# Ensure the 'Sample' column exists and is consistent 
bin_count <- bin_count %>% rename(Sample = Sample)
nobin_count <- nobin_count %>% rename(Sample = Sample)

# Full join to combine counts 
combined_counts <- dplyr::full_join(bin_count, nobin_count, by = "Sample") %>%
  tidyr::replace_na(list(Unique_BINs = 0, Unique_Unknown_Sequences = 0))

# Ensure all samples from sample.w.data.list are included
missing_samples <- setdiff(sample.w.data.list, combined_counts$Sample)
if(length(missing_samples) > 0){
  combined_counts <- bind_rows(
    combined_counts,
    data.frame(
      Sample = missing_samples,
      Unique_BINs = 0,
      Unique_Unknown_Sequences = 0
    )
  )
}

# Get finite positive values for log scale
vals <- c(combined_counts$Unique_Unknown_Sequences,
          combined_counts$Unique_BINs)

# log10 scale cannot use zero or negative numbers
vals <- vals[is.finite(vals) & vals > 0]

# symmetric square bounds
plot_min <- floor(log10(min(vals)))
plot_max <- ceiling(log10(max(vals)))

# Plot
scatter_plot <- combined_counts %>%
  ggplot(aes(x = Unique_Unknown_Sequences, y = Unique_BINs,
             text = paste0("Sample: ", Sample,
                           "<br>ASVs with BIN match: ", Unique_BINs,
                           "<br>ASVs without BIN match: ", Unique_Unknown_Sequences))) +
  geom_point(color = "#1EC2A0", size = 3, alpha = 0.7) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50") +
  scale_x_log10(limits = c(10^plot_min, 10^plot_max)) +
  scale_y_log10(limits = c(10^plot_min, 10^plot_max)) +
  coord_equal() +          # ensures 1:1 aspect ratio
  theme_bw(base_size = 14) +
  labs(x = "# ASVs without BIN match (log10)",
       y = "# ASVs with BIN match (log10)")

ggplotly(scatter_plot, tooltip = "text")


```

# BIN richness (extrapolation based on 2X observations) for all `r sample.nos.w.data` samples {.section-box .text-center}

```{r}
#| warning: false
#| fig-width: 8
#| fig-height: 7

# Data for the accumulation curves
records_BIN<-metabarcode_data_by_sample%>%
  filter(`BIN Match Status`=='BIN_MATCH')%>%  
  group_by(`BIN Hit`)%>%
  dplyr::summarise(records = n())%>%
  pull(records)

# Give name to the output 
names(records_BIN) <- "BIN estimate"

# iNEXT analysis
out_abund <- iNEXT(records_BIN,
                   datatype = "abundance", 
                   q = 0)

# Extract observed value for plot
obs_value<-out_abund$iNextEst$size_based%>%
  select(m,Method,qD)%>%
  filter(Method=='Observed')

# Extract observed value for plot
extrapolated_value<-tail(out_abund$iNextEst$size_based,1)%>%
  select(m,Method,qD)

# Generate plot
options(scipen = 999) # Convert scientific notation to numbers

div_plot<-ggiNEXT(out_abund,
        se=T,
        type=1,
        color.var = "Assemblage") +
  scale_color_manual(values = c('#CC4945'))+
  theme_bw(base_size = 14)+
  theme(axis.text.x = element_text(colour = "#011826", 
                                   hjust = 1),
        panel.grid.minor.x = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.y = element_blank(),
        legend.position = 'none')+
   geom_hline(yintercept = round(obs_value$qD,0), 
             linetype = "dashed", 
             color = "#011826")+
  annotate(
    "text",
    x = round(obs_value$qD,0)+round(obs_value$qD,0)*0.1,       
    y = round(obs_value$qD,0)+round(obs_value$qD,0)*0.045,      
    label = paste("Observed BIN richness: ",round(obs_value$qD,0),sep=''),  
    color = "#011826",
    vjust = -0.6    
  )+
  geom_hline(yintercept = round(extrapolated_value$qD,0), 
             linetype = "dashed", 
             color = "#CC4945")+
  annotate("text",
    x = round(extrapolated_value$qD,0)+round(extrapolated_value$qD,0)*0.1,       
    y = round(extrapolated_value$qD,0)+round(extrapolated_value$qD,0)*0.025,      
    label = paste("Extrapolated BIN richness: ",round(extrapolated_value$qD,0),sep=''), 
    color = "#011826",
    vjust = -0.6)+    
  labs(title = "",
    x = "Number of occurrences",
    y = "BIN richness") 
  

ggplotly(div_plot)
```

```{r}
#| warning: false
#| echo: false
#| results: 'asis'
#| column: page

cat(paste0('
<footer style="margin-top: 3em; padding-top: 1em; border-top: 1px solid #011826; text-align: center; font-size: 0.9em; color: #011826;">
  <p> Centre for Biodiversity Genomics.<br>
  Created: ', format(Sys.Date(), "%B %d, %Y"), '</p>
</footer>
'))
```
